---
title: "Mayo Clinic Yelp Review Text Analysis"
author: RexManglicmot
output: 
  github_document: 
    toc: yes

always_allow_html: true
---

## Introduction
<p align="center"> <img src="https://target.scene7.com/is/image/Target/GUEST_1ec18fe8-25da-4ae9-a754-33f87564fdcb?wid=725&hei=725&qlt=80&fmt=webp" width="400"/> </p>
What is the best hospital in the United States? Which one is closest to me? What do real patients have to say about a particular hospital? These are just a few of the many questions Americans face as they navigate the complex U.S. healthcare system.

In 2022, average healthcare spending per person in the U.S. reached $12,530—up from $11,462 in 2019.^[https://www.chcf.org/publication/2022-edition-health-care-costs-101/] With 6,093 hospitals operating across the country,^[https://www.aha.org/statistics/fast-facts-us-hospitals] choosing where to receive care is both a personal and financial decision. Unlike single-payer systems in countries like Canada, the United Kingdom, or Australia, the U.S. healthcare system is a mix of private, public, and nonprofit institutions. Consequently, not all hospitals are created equal—some are well-resourced and highly regarded, while others struggle with quality and access.

To help guide patients, U.S. News & World Report publishes an annual ranking of the best hospitals in the United States.^[https://health.usnews.com/health-care/best-hospitals/articles/faq-how-and-why-we-rank-and-rate-hospitals] While the list evaluates about 4,500 facilities, it does not include every hospital in the country. For the 2022–23 rankings, the Mayo Clinic in Rochester, Minnesota, was named the #1 hospital.

To gain further insight into what makes the Mayo Clinic exceptional—and to complement traditional clinical and administrative metrics—this project applies a text analysis approach to patient reviews on Yelp. Yelp is a widely used platform for reviewing businesses, including healthcare institutions, and offers a unique window into the lived experiences of patients.

Objectives
This project has three main goals:

Identify recurring word patterns and themes that reflect authentic patient sentiment in a reproducible way.

Provide actionable insights for hospital administrators to better understand the patient experience.

Offer benchmarks for lower-ranked hospitals to reflect on how patient-centered services might be improved.

To address these objectives, this report applies a suite of text mining methods to analyze unstructured Yelp reviews of the Mayo Clinic.

Report Structure:
This project is organized into the following sections:
1. Webscraping Yelp Data
2. Loading the Libraries
3. Loading the Data
4. Cleaning the Data
5. Exploratory Data Analysis
6. Column Charts and WordClouds
7. Sentiment: Positive and Negative Words
8. Limitations
9. Conclusion
10. Appendix
11. Inspiration for This Project



## Webscraping Yelp Data
Yelp data was scrapped on via the [Yelp](https://www.yelp.com/) website. Within the search engine bar, I typed in *Mayo Clinic* and used the first "business result to scrape the data as it's location, Rochester Minnesota matched on US News. Further, the business had a "blue-checkmark" with "Claimed" indicating that the business was the legitimate Mayo Clinic.

There were 228 reviews in total and the goal was to scrape all 228 reviews containing these 4 metrics:

1. reviewer name
2. reviewer location
3. review rating
4. review text

(Photos were excluded from scraping due to the focus on text-based analysis.)

After several hours of debugging and inspecting HTML tags, I successfully extracted all 228 reviews. One key challenge was avoiding the inclusion of business replies, such as official Mayo Clinic responses, which could have inflated the review count. Adjusting the scraping logic helped prevent this.

A special acknowledgement to Samer Jijjazi for his YouTube tutorial in learning how to scrape data from Yelp.^[https://www.youtube.com/watch?v=qyGYItbMKkM]. Watching his videos greatly helped me get the data needed to webscrape.

The code used for scraping is included below. The resulting dataset (Mayo_Clinic.csv) is also available in this repository.
```{r Webscraping Yelp data, eval=FALSE}

#load libraries
library(tidyverse)
library(rvest)

#create an object to store the webpage address
url <- 'https://www.yelp.com/biz/mayo-clinic-rochester-12?osq=Mayo+clinic'

#convert the url to an html object for R processing
webpage <- read_html(url)

#create object to know page number on the webpage
webpageNum <- webpage %>%
  html_elements(xpath = "//div[@class= ' border-color--default__09f24__NPAKY text-align--center__09f24__fYBGO']") %>%
  html_text() %>%
  str_extract('of.*') %>%
  str_remove('of ') %>%
  as.numeric()

#create a sequence to iterate for page number
webpageSeq <- seq(from = 0, to = (webpageNum * 10)-10, by = 10)

#store items into empty objects
reviewer_name_all = c()
reviewer_location_all = c()
review_rating_all = c()
review_text_all = c()

#create a for loop to get values throughout the 23 pages
for (i in webpageSeq) {
  #need to create an if statement because the 1st page web address
  if (i == 0) {
    webpage <- read_html(url)
    #need to create else because webpage has more content than 1st page web 
    #address
  } else {
    webpage <- read_html(paste0(url, '&start=', i))
  }
  
  #reviewer name
  reviewer_name <- webpage %>%
    #return elements that I specify via xpath method
    #xpath is useful for locating elements
    #// means to search within entire document
    #* means to return any of the elements
    # div means to search within document with the div tags
    html_elements(xpath = "//div[starts-with(@class,' user-passport')]") %>%
    #look within a tag within the previous element
    html_elements(xpath = ".//a[starts-with(@href, '/user_details')]") %>%
    html_text()
  
  #reviewer location
  reviewer_location <- webpage %>%
    #location is within the same div tag, so use same code
    html_elements(xpath = "//div[starts-with(@class,' user-passport')]") %>%
    #location is also located within the span tag
    html_elements(xpath = ".//span[@class= ' css-qgunke']") %>%
    html_text() %>%
    #remove "Location" 
    #pipe remaining values that are not "Location"
    .[. !='Location']
  
  #review rating
  review_rating <- webpage %>%
    html_elements(xpath = "//div[starts-with(@class, ' review')]") %>%
    #within div tag there is an aria-label
    #contains function to look for aria-label that has rating  
    html_elements(xpath = "(.//div[contains(@aria-label, 'star rating')])[1]")%>%
    #ratings are not text, so must use different method and specify which 
    #attribute to obtain
    html_attr('aria-label') %>%
    #remove star rating
    str_remove_all(' star rating') %>%
    #convert into a numeric
    as.numeric()
  
  #review text
  review_text <- webpage %>%
    html_elements(xpath = "//div[starts-with(@class, ' review')]") %>%
    #look throughout webpage with the p tag
    #to get the first comment and not worr about business comment,
    #need to put in brackets
    html_elements(xpath = "(.//p[starts-with(@class, 'comment')])[1]") %>%
    #html_elements(xpath = ".//span[starts-with(@class, ' raw')]") %>%
    html_text()
  
  #appending to appropriate objects
  reviewer_name_all = append(reviewer_name_all, reviewer_name)
  reviewer_location_all = append(reviewer_location_all, reviewer_location)
  review_rating_all = append(review_rating_all, review_rating)
  review_text_all = append(review_text_all, review_text)
  
}

#create a dataframe containing appended values
Mayo_Clinic <- data.frame('name' = reviewer_name_all,
                          'location' = reviewer_location_all,
                          'rating' = review_rating_all,
                          'text'= review_text_all)

#view csv file
#head(Mayo_Clinic)
```

## Loading the Libraries

```{r, message=FALSE, EDA }
#load libraries
#install.packages('tidyverse') #had to re-install for some reason on 11/12/22
#install.packages('ggraph') #had to re-install for some reason on 11/12/22
#install.packages('DT')
#install.packages('gt')

library(tidyverse)
library(tidytext)
library(widyr)
library(RColorBrewer)
library(wordcloud)
library(igraph)
library(ggraph)
library(gt)
library(dplyr)
```

## Loading the Data
```{r, message=FALSE}
#load Mayo Clinic data
data <- read.csv('Mayo_Clinic.csv')
```

## Cleaning the Data
Since the dataset has been successfully loaded, we'll begin by exploring its structure. Instead of using the basic str() function, I’ll expand my R skills by applying other functions to examine the variable names and dimensions of the dataset.
```{r}
#get the names of the variables
colnames(data)

#get the number of rows in the dataset
dim(data)
```
From the output, we can see that there is an extra column named X, which is simply an index column generated when the CSV file was saved. This column is not needed for our analysis, so we’ll remove it.

```{r}
#remove the X column
data <- data %>%
  select(-X)

#get the names other than the colnames funciton
names(data)
dim(data)
```

Now that the X column has been removed, we can confirm that the dataset contains only the relevant columns: reviewer name, location, rating, and review text.
```{r}
#View first rows of the observations of the data
head(data)

#how many observations are in the text column
length(data$text)
```
At first glance, we see that the text column appears empty for some early observations, except for the fifth entry. However, the total number of observations in the text column confirms that all 228 reviews are present.

When viewing the data frame in RStudio, the text appears right-aligned by default. This alignment is purely a visual artifact of how R prints character vectors in console views and does not impact the content or structure of the data itself.

Great, now let's see if there are duplicates AND if there is more than one reviewer in the dataset. 
```{r}
#check for duplicates
sum(duplicated(data))

#quick check to see if reviewer_name appears more than once
data %>%
  count(name, sort = TRUE) %>%
  #let's arbitarailys see the first 10 rows as base comparison
  head(n=10)
```
The results show that there are no duplicate rows in the dataset. However, five reviewers submitted two reviews each, meaning 5 individuals account for 10 reviews. This brings the count of unique reviewers to 223 out of the 228 total reviews.

This step confirms that while the dataset is nearly one-to-one between reviewer and review, there are a few repeat reviewers worth noting.

In cleaning our data, let's see if there are any NA values and which columns. 
```{r}
#if any NA, then where in the dataset
which(is.na(data))

#quick check to see and sum if any NA are in rating variable
sum(is.na(data))

```

The output confirms that there are no missing values (NA) in the dataset. All rows across all columns—reviewer name, location, rating, and text—are complete and ready for analysis.

As a final check before beginning analysis, we verify that all values in the rating column fall within the expected range of 1 to 5. We also ensure there are no unexpected characters or text entries in the column.
```{r}
#quick check to see if below 1 OR above 5
any(data$rating > 5 | data$rating <1)

#create functions
numbers_only <- function(x) !grepl("\\D", x)
letters_only <- function(x) !grepl("[^A-Za-z]", x)

#let's double check the ratings column
sum(numbers_only(data$rating))
sum(letters_only(data$rating))

```

The results confirm that 1) No ratings are outside the 1–5 range 2) All 228 entries are numeric and 3) There are no alphabetic or invalid characters in the rating column.

Now that the dataset has been thoroughly validated and cleaned, we’re ready to begin exploring the data.

## Exploratory Data Analysis

To begin exploring the dataset, we first examine the distribution of patient ratings using summary statistics.
```{r}
#calculate summary statistics
summary(data$rating)
```

From the output, we observe that the median and maximum rating are both 5, while the mean is approximately 3.39. This suggests that a large number of patients gave Mayo Clinic the highest rating, indicating a positively skewed distribution.

The best way to visualize this pattern is with a bar chart, which we'll generate in the next step.

```{r}
#create a barchart to count the values
ggplot(data, aes(x=rating)) +
  geom_bar(color='black', fill='steelblue', width = .7) +
  scale_fill_brewer(palette="Dark2") +
  theme_minimal() +
  labs(title = 'Yelp Review Ratings',
        x = 'Star Ratings',
        y = 'Count')
```

The bar chart clearly shows a polarized distribution in Yelp reviews for the Mayo Clinic. The majority of reviews are either 1-star or 5-star, with significantly fewer reviews in the middle categories (2, 3, and 4 stars).

This pattern suggests that patient experiences tend to be highly positive or highly negative, rather than moderate. These extremes in sentiment indicate that reviewers may be motivated to leave feedback only when their experiences are particularly memorable—either very good or very poor.

Let's get the actual values of these counts by building a table.
```{r}
#create dataframe to build stacked barchart
rating <- c(1,2,3,4,5)
total <- c((sum(data$rating == 1)),
           (sum(data$rating == 2)),
          (sum(data$rating == 3)),
          (sum(data$rating == 4)),
          (sum(data$rating == 5)))

#create a table of all the rating sums
table3 <-data.frame(rating, total)
print(table3)
```
From the table above, we see that 1) 72 reviews gave a rating of 1 star and 2) 121 reviews gave a rating of 5 stars. Together, those two categories account for nearly 85% of all reviews. This reinforces the earlier insight that most patients rated their experience at the Mayo Clinic as either very positive or very negative, with relatively few moderate experiences in between.

WORK ON!!!!
START
What we can do for the next steps are to separate the 228 reviews into 3 sections; 1 rating, 5 rating, and overall rating and see what words are associated within each. That would be interesting to figure out. 
STOP

## WordClouds
In this project, wordclouds will be used to illustrate common words patients use when reviewing the Mayo Clinic. These visualizations help surface emotionally charged, repeated, or thematically important words without requiring the reader to sift through hundreds of individual reviews. Wordclouds emphasize the most frequently used words in a body of text by enlarging them proportionally to their frequency, making them ideal for quickly identifying key themes or repeated language.

In order to create a wordcloud, we need to break each review into individual words and represent each word as a separate observation. This transformation allows us to count word frequency across all reviews. As a result, the total number of observations will increase significantly from the original 228 rows.

This process is known as tokenization—splitting sentences into their component words. To begin, we remove unnecessary columns (name, location, and rating) and store the cleaned text data in a new object, data2. We also verify that the text column is a character type before proceeding.
```{r}
#Tokenize data

#drop irrelevant columns by indexing 
data2 <- data[-c(1:3)] 

#let's double check the str to see it is a character 
is.character(data2$text)
```
Output [1] TRUE confirms that the column is suitable for tokenization.

Now let's break up the sentences into individual words (tokenizing).
```{r}
#tokenize data
data3<- data2 %>%
  #break up each sentence in the text and assign them an individual row
  unnest_tokens(word, text)

#count token data
data3%>%
  count(word) %>%
  #count with the highest tally of words on top of the list
  arrange(desc(n)) %>%
  head(n=10)
```

Because the meaning of an English sentence is shaped by the combination of its words, it's important to identify and remove those that do not contribute significant meaning—often referred to as “stop words” or “filler words” (e.g., "the," "a," "an," "of," etc.). These words are extremely common but do not help us understand the content or sentiment of a review. To create a more informative and meaningful wordcloud, we’ll clean the tokenized data by removing these stop words using a predefined dictionary. We’ll perform this step using an anti-join and save the cleaned data into a new object called data4.

```{r, message=FALSE}
#takeout stop words
data4 <- data2 %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) #we have less words now, like 1/3 got erased
```


```{r}
#count token data again without the stop words
data4 %>%
  count(word) %>%
  arrange(desc(n)) %>%
  head(n=10)
```
We observe that the words "mayo" and "clinic" are the most frequently used terms, ranking 1st and 2nd, respectively. To further refine our analysis and generate a more insightful wordcloud, we should consider adding these words to the stop word dictionary and removing them from the dataset.

```{r}
custom_stop_words <- tribble(
  ~word, ~lexicon,
  'mayo', 'CUSTOM',
  'clinic', 'CUSTOM',
)

stop_words2 <- stop_words %>%
  bind_rows(custom_stop_words)
```
Let's see if that works

```{r, message=FALSE}
#takeout out custom words
data5 <- data4 %>%
  anti_join(stop_words2)
```


```{r}
#count token data again without the stop words and arrange the 
data5 %>%
  count(word) %>%
  arrange(desc(n)) %>%
  head(n=15)
```
It worked. So, lets get rid of numbers in the list.
```{r}
#not interested in numbers, need to filter out
data6 <- data5 %>%
  #include words that has at least one alphabetical charachter in each out
  filter(str_detect(word, '[:alpha:]')) 
```
Now let's get arrange the by the number of the most frequent words and get rid of words that occur **once**, **twice**, and **thrice** since we are only interested in the most frequent words.

```{r}
#count token data again without the numbers
data7 <- data6 %>%
  count(word) %>%
  arrange(desc(n)) %>%
  filter(n!= 1 & n!= 2 & n!=3)

dim(data7)
str(data7)

```
After a series of preprocessing steps—including tokenization, removal of stop words, filtering out non-alphabetic tokens, and excluding infrequent terms—we arrive at a clean dataset consisting of 659 unique words that appeared at least four times in the corpus. These remaining words, such as “care,” “time,” and “doctor,” reflect the most meaningful and frequently mentioned concepts in the reviews. This cleaned data provides a more accurate linguistic representation of the core themes discussed by patients, removing filler or non-informative content. With this refined dataset, we are now well-positioned to generate insightful visualizations, such as word clouds or bar charts, to better understand the topics and concerns most commonly raised in relation to Mayo Clinic.

```{r}
library(ggplot2)

#lets filter for top 50 words
data8 <- data7 %>%
  filter(n > 50) %>%
  #rearrange for ggplot 
  mutate(word2 = fct_reorder(word, n))

#plot the data
ggplot(data8, aes(x=word2, y= n)) +
  geom_col(fill = 'lightblue', color= 'black') + 
  coord_flip() +
  labs(title = "Top Words That Patients Use To Describe the Mayo Clinic ",
       x = 'Count',
       y = 'Words') +
  theme_classic()
```

```{r, warning=FALSE}
wordcloud(
  words = data7$word,
  freq = data7$n,
  max.words = 50,
  colors = 'lightblue'
)
```

The wordcloud above shows the overall patient word count from the 228 reviews. Now let's create two additional wordclouds for those who rated Mayo Clinic 1 and 5 stars. 

```{r, message=FALSE}
#take original data set and pipe
data_A <-data %>%
  # get rid of all the columns name and location
  select(-name, -location) %>%
  # get rid of ratings 2, 3, and 4
  subset(rating != 2 & rating !=3 & rating !=4) %>%
  arrange(rating)

#create object that contains only 1 and 5 ratings only
data_A1 <- data_A %>%
  subset(rating != 5)

data_A5 <- data_A %>%
  subset(rating != 1)

#follow tokenize process for 1 rating
data_A1T <- data_A1 %>%
  #get rid of the rating category as this is not important anymore
  select(-rating) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words2) %>%
  filter(str_detect(word, '[:alpha:]')) %>%
  count(word) %>%
  arrange(desc(n)) %>%
    filter(n!= 1 & n!= 2 & n!=3)

#follow tokenize process for 5 rating
data_A5T <- data_A5 %>%
  #get rid of the rating category as this is not important anymore
  select(-rating) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words2) %>%
  filter(str_detect(word, '[:alpha:]')) %>%
  count(word) %>%
  arrange(desc(n)) %>%
  filter(n!= 1 & n!= 2 & n!=3)
```

```{r}
wordcloud(
  words = data_A1T$word,
  freq = data_A1T$n,
  max.words = 50,
  colors = 'red')
  
  wordcloud(
  words = data_A5T$word,
  freq = data_A5T$n,
  max.words = 50,
  colors = 'orange')
```

Interesting. There is not much of a big difference amongst the three groups. There are two insights from this:

1. Maybe further filtering of words should be the next steps or, 
2. Wordclouds are not the best medium to understand a patient views on a healthcare system.

Next, lets use sentiment dictionaries on our three groups; overall, 1 rating, and 5 rating.

## Sentiment: Positive and Negative Words

Sentiment Analysis (SA) is a method to evaluate written text or language based on a dictionary. In other words, SA is used to extract meaning from a text. SA provides the ability to understand an observation's (i.e., customer) experience regarding a topic. SA is used when there is large volume of text and to create a generalization of such.

### Appplications

* Customer feedback. Understanding user review comments on a web platform (i.e., Twitter)
* Analyze survey responses
* Competitor research

### Pros

* Diminish the risk of bias from a person investigating comments. As such, it diminish personal perceptions of the data and takes away the person's mood and other incalculable biases.
* Quick to implement.


### Cons

```{r, message=FALSE}
#install.packages('SentimentAnalysis') #installed on 12/12/22
# install.packages('textdata') #installed on 12/12/22

#load libraries
library(SentimentAnalysis)
library(textdata)

#use data6 as it is the most funnelled data
sentiment_data <- data6 %>%
  #used the nrc dictionary
  inner_join(get_sentiments('nrc'))

sentiment_data_nrc <- sentiment_data %>%
  count(sentiment) %>%
  mutate(sentiment2 = fct_reorder(sentiment, n))

ggplot(sentiment_data_nrc, aes(x= sentiment2, y=n)) +
  geom_col(fill = '#669933') +
  coord_flip() +
  theme_classic() +
  labs(title = 'Mayo Clinic Sentiment Counts in NRC Dictionary',
       x= 'Sentiment',
       y = 'Counts')

```

Interesting.

Insights:

* Based on the NRC dictionary, majority of the patients expressed the sentiment "positive" in their reviews, which had the most counts, followed by trust. 
* The following four sentiments thereafter, as a group, be categorized as ambiguous and negative.

Interesting. What does this tell us? There is a mix big of emotions from patients. It's best to try another dictionary. So, let's try 'Loughran'

```{r, message=FALSE, warning=FALSE}
#use data6 as it is the most funneled data
sentiment_data2 <- data6 %>%
  #used the nrc dictionary
  inner_join(get_sentiments('loughran'))

sentiment_data_loughran <- sentiment_data2 %>%
  count(sentiment) %>%
  mutate(sentiment2 = fct_reorder(sentiment, n))

ggplot(sentiment_data_loughran, aes(x= sentiment2, y=n)) +
  geom_col(fill = '#FFCC66', width = 0.6) +
  coord_flip() +
  theme_classic() +
  labs(title = 'Mayo Clinic Sentiment Counts in Loughran Dictionary',
       x= 'Sentiment',
       y = 'Counts')
```

Even more interesting.

Insights:

* Using this dictionary, we see that many of the reviews a categorized as negative.
* There is a 2x count of negative vs positive reviews. 
* Uncertainty, litigious, constraining categories could further be included as negative.

## Limitations
Depending on the dictionary, there could be a diverse set of insights from patients reviews. Using only two dictionaries, we see that the reviews can easily be interpreted in different ways. Thus, sentiment analysis although a good tool to understand patient perceptions are not alone should be a sole tool to understand. 

## Conclusions

## Appendix


## Inspiration for this project

Having worked at a non-profit hospital for six years made me realize the importance of healthcare providers, patients, and the entire health ecosystem. I wanted to see how my hospital ranked among others. Needless to say, we did not make the rankings. That realization did spark an interest to see what metrics did the top hospitals had over others. I wanted to take a different approach and use patients reviews as a metric. 

Using patient reviews provides an unadultered way to understand patients' perspectives. 


